Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-06-15 Manuscript/manuscript.tex
saving work

The hCI methods introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space defined by the reference and the value of $h$.
{\QP} \cite{Garniron_2019} was employed to perform all the CI calculations presented here.
In a given calculation, the energies are considered to be converged when the (largest) Epstein-Nesbet second-order perturbation correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
%by spanning the most important regions of the truncated Hilbert space
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delievering fairly converged absolute and excitation energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}

@@

% rPT2 vs PT2

We further assesssed the effect of the EN2 perturbative correction for both the open-shell systems studied here 
and the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) addressed in our first report on hCI. \cite{Kossoski_2022}
The full set of PECSs is shown in the {\SupInf}.
At stretched geometries, the EN2 correction can lead to kinks and large jumps in the PECSs.
Such problems were encountered at some CI methods for \ce{CN}, vinyl, \ce{H8}, and \ce{H7}, and to a less extent, for \ce{N2} (only for hCI1).
For the remaining five systems, the EN2 correction did not produce problematic PECs.

Around the equilibrium geometry, the EN2 correction produced well-behaved PECs at all levels of CI and for all systems, 
except for \ce{CN}, which is related to its different states around equilibrium.
%
xe
\ce{OH} massive improvement
\ce{CN} \ce{H4} \ce{N2} \ce{F2} \ce{HF} big improvement
\ce{H7} \ce{H8} vinyl ethylene some improvement
%
freq
\ce{OH} massive improvement
\ce{HF} \ce{H8} vinyl ethylene some improvement
\ce{CN} \ce{F2} \ce{N2} \ce{H4} \ce{H7} big improvement

@@

In Tab.~\ref{tab:1} we present the statistical errors of the different methods, organized by their computational scaling.
We found that $\Delta$hCI2 and $\Delta$CISD display comparable MAEs, despite the somewhat more favourable MSE and RMSE of the latter.
$\Delta$hCI2 still tends to underestimate the excitation energies, though to a much lesser extent than observed for the lower-order hCI methods.

@@

Despite the rather poor performance of lower-order hCI methods for excitation energies,
we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
For that, we performed calculations for a subset of 14 excited states of small molecular systems (\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
with corresponding statistical errors shown in Tab.~\ref{tab:2}.
Even though this is a small set, it should be enough to obtain the main trends on the statistics of the different methods.

@@

The case of $\Delta$hCI2 was discussed above for the larger set of excitation energies.
Here we further notice that hCI2 was found to be somewhat more accurate than CISD, even though the errors are still too large.
However, this improvement does not reflect on the state specific route (compare $\Delta$hCI2 and $\Delta$CISD in Tabs.~\ref{tab:1} and ~\ref{tab:2}).
Moving one step further, even though the state-specific route ($\Delta$hCI2.5) is far more accurate than the ground-state based route (hCI2.5),
it is actually less accurate then $\Delta$hCI2.
An improved performance is found for the methods displaying an $\order*{N^8}$ computational scaling.
The four methods, hCI3, $\Delta$hCI3, CISDT, and $\Delta$CISDT, display quite similar MAEs and RMSEs.
Therefore, there is no advantage by moving from excitation-based to hierarchy-based CI at this order, just as discussed above for the $\Delta$hCI2 and $\Delta$CISD.
Similarly, the $\Delta$hCI3 does not improve with respect to hCI3, as also observed for $\Delta$CISDT and CISDT. \cite{Kossoski_2023}
We notice, however, the negative MSEs for the state-specific methods, in contrast to the positives values obtained for the ground-state based route.
At the next order, $\Delta$hCI3.5 has significantly smaller errors, but at a costly $\order*{N^9}$ scaling.

@@

We refer back to Tab.~\ref{tab:1} to discuss the performance of seniority-based CI for excited states.
sCI2 delivers quite poor results (MAE of \SI{1.06}{eV}), whereas $\Delta$sCI2 reduces the statistical errors (MAE of \SI{0.60}{eV}).
This improvement, however, is not as expressive as typically observed for excitataion-based and hierarchy-based methods.
(Compare, for instance, with the cases of hCI2.5 and $\Delta$hCI2.5 shown in Tab.~\ref{tab:2}).
The large errors, even at the state-specific route, combined with the formal $\order*{e^N}$ scaling of these methods,
render them unsuitable for calculations of accurate excitation energies.

@@

We conclude that 
Meanwhile, 
, and therefore not useful.

>>>>

The hCI methods introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space defined by the reference and the value of $h$.
{\QP} \cite{Garniron_2019} was employed to perform all the CI calculations presented here.
In a given calculation, the energies are considered to be converged when the (largest) renormalized second-order Epstein-Nesbet (EN2) perturbation correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
%by spanning the most important regions of the truncated Hilbert space
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delievering fairly converged absolute and excitation energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}

@@

We further assesssed the effect of the EN2 perturbative correction for both the open-shell systems studied here 
and the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) addressed in our first report on hCI. \cite{Kossoski_2022}
The full set of PECSs is shown in the {\SupInf}.
At stretched geometries, the perturbative correction can lead to kinks and large jumps in the PECs.
Such problems were encountered at some levels of CI for \ce{CN}, vinyl, \ce{H8}, and \ce{H7}, and to a less extent, for \ce{N2} (only for hCI1).
For the remaining five systems (\ce{HF}, \ce{F2}, ethylene, \ce{H4}, and \ce{OH}),
the EN2 correction produced smooth PECs at dissociation
Around the equilibrium geometry, the perturbative correction produced well-behaved PECs at all levels of CI and for all systems, except for \ce{CN},
which is related to the close-lying states around equilibrium.

We found that the EN2 perturbative correction generally reduces the errors of all observables considered here, for a given level of excitation-based CI and hCI routes.
As such, it accelerates the rate of convergence as a function of the excitation degree or hierarchy parameter.
Importantly, the perturbative corrected hCI route still outperforms the excitation-based counterpart, just as for the uncorrected comparison.
Furthermore, the reduction of the errors is overall more monotonic than without the perturbative correction.
This is illustrated in Fig.\ref{} for the particular case of \ce{F2}.
For the sCI route, however, the EN2 correction does not bring a clear improvement to render it competitive in relation to the excitation and hierarchy-based alternatives.

% rPT2 vs PT2 there are important differences

%xe
%\ce{OH} massive improvement
%\ce{CN} \ce{H4} \ce{N2} \ce{F2} \ce{HF} big improvement
%\ce{H7} \ce{H8} vinyl ethylene some improvement

%freq
%\ce{OH} massive improvement
%\ce{CN} \ce{F2} \ce{N2} \ce{H4} \ce{H7} big improvement
%\ce{HF} \ce{H8} vinyl ethylene some improvement

@@

In Tab.~\ref{tab:1} we present the statistical errors of the different methods, organized by their computational scaling.
% hCI1 and hCI1.5
We found that $\Delta$hCI2 and $\Delta$CISD display comparable MAEs, despite the somewhat more favourable MSE and RMSE of the latter.
This means that the additional determinants present in the former method recover a larger fraction of the correlation energy for excited state than for the ground state,
thus creating a bias toward underestimated excitation energies.
Even though this unbalance is still observed for $\Delta$hCI2, it is much less serious than observed for the lower-order state-specific hCI methods.

@@

Despite the rather poor performance of lower-order hCI methods for excitation energies,
we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
For that, we performed calculations for a subset of 14 excited states of small molecular systems (\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
with corresponding statistical errors shown in Tab.~\ref{tab:2}.
Even though this is a small set, it should be enough to obtain the main trends on the accuracy of higher-order methods.

@@

The case of $\Delta$hCI2 was discussed above for the larger set of excitation energies.
Here we further notice that hCI2 was found to be somewhat more accurate than CISD, even though the errors are still too large.
However, this improvement does not reflect on the state specific route (compare $\Delta$hCI2 and $\Delta$CISD in Tabs.~\ref{tab:1} and ~\ref{tab:2}).
Moving one step further, even though the state-specific route ($\Delta$hCI2.5) is far more accurate than the ground-state based route (hCI2.5),
it is actually less accurate then $\Delta$hCI2.
An improved performance is found for the methods displaying an $\order*{N^8}$ computational scaling.
The four methods, hCI3, $\Delta$hCI3, CISDT, and $\Delta$CISDT, present quite similar MAEs and RMSEs.
Therefore, there is no real advantage in moving from excitation-based to hierarchy-based CI at this order, just as discussed above for the $\Delta$hCI2 and $\Delta$CISD.
Similarly, the $\Delta$hCI3 is not more accurate than hCI3, as also observed for $\Delta$CISDT and CISDT. \cite{Kossoski_2023}
We notice, however, that the state-specific methods present negative MSEs, in contrast to the positives values obtained in the ground-state based route.
This reflects an umbalanced description of correlation effects, favoring the excited states.
At the next order, $\Delta$hCI3.5 has significantly smaller errors, but at a costly $\order*{N^9}$ scaling.

@@

We refer back to Tab.~\ref{tab:1} to discuss the performance of seniority-based CI for excited states.
sCI2 delivers quite poor results (MAE of \SI{1.06}{eV}), whereas $\Delta$sCI2 reduces the statistical errors (MAE of \SI{0.60}{eV}).
This improvement, however, is not as expressive as observed for excitataion-based and hierarchy-based methods.
Compare, for instance, with the cases of hCI2.5 and $\Delta$hCI2.5 shown in Tab.~\ref{tab:2}.
The large errors, even following the state-specific approach, combined with the formal $\order*{e^N}$ scaling of these methods,
render seniority-based CI methods unsuitable for calculations of excitation energies.

@@

%three CI routes (excitation, hierarchy, and seniority-based).

We have considered both closed and open shell systems, considering ground-state properties and excitation energies, for a diverse set of molecular systems.
Extending our previous findings for closed-shell \cite{Kossoski_2022} to open-shell systems,
we conclude that the hCI route has an edge in comparison to excitation-based CI,
and that seniority-based CI methods lead to far less accurate results than the two former alternatives, for a given computational cost.
We also found that, for closed- and open-shell systems,
%EN2 perturbative corrections lead to faster and more monotonic convergence of different properties
EN2 perturbative corrections lead to substantial improvements for hCI and excitation-based CI routes (while keeping the upper hand for the former), but do not help much with the deficiencies of the sCI route.

We further assessed the performance of sCI and hCI methods, based on either HF ground-state orbitals or state-specific orbitals, to describe excited states of closed-shell systems and radicals.
Overall, sCI methods present relatively poor results considering their high computational costs, and thus should not be useful in practice.
%it is fair to conclude that
